Open-source tools in R for landscape ecology

### 2021-01-27 paper/paper.tex
Adding simple features and adding some reference currently missing

The growing body of \emph{R} packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.
Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g.~Python) for landscape ecology exist, in this review we focus on software implemented in the \emph{R} programming language.
For more general overviews of open-source software for landscape ecology see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
The \emph{R} programming language can be a very powerful tool. In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the \emph{R} environment (Fig. \ref{fig:fig-workflow}).

@@

The guaranteed technical quality standard on \emph{CRAN} requires more initial work for developers compared to \emph{GitHub}, while it ensures for users that the package can be installed on their machine.
Additionally, the technical quality standard on \emph{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \emph{R} packages, i.e., package \emph{x} requires and uses code from package \emph{y} .
CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability and reproducibility of code.
Thus, most packages can be found on both platforms, and many developers use \emph{GitHub} for regular development and \emph{CRAN} to publish stable releases of the packages.
Furthermore, online communities like \emph{rOpenSci} also provide a peer-review process for code quality.
However, while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
Packages that deal with such issue include \emph{groundhog} \cite{Simonsohn2021}, \emph{packrat} \cite{Ushey2018}, or \emph{renv} \cite{Ushey2020}.
For more information about \emph{R} package development in general, see \cite{Wickham2015}.

@@

The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands .
The \emph{sp} package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 \emph{R} packages directly depend or imports \emph{sp}.
However, \emph{sp} is not actively developed anymore, and its recommended successor is the \emph{sf} package \cite{Pebesma2018}.
Besides many advantages and strength of the \emph{sf} package in terms of spatial data handling, it also integrates into the widely used \emph{tidyverse} packages \cite{Wickham2019a}.

@@

Nowadays, spatial data for many scales is available from an abundance of online-accessible sources.
A lot of this data are publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an \emph{R} session.
Since publicly available data is becoming more prominent, so are \emph{R} packages to access them.
Packages include \emph{rnaturalearth} \cite{South2017} to access the Natural Earth database to download region and country data, the \emph{elevatr} package to access raster elevation data \cite{R-elevatr}, the \emph{rgbif} package to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the \emph{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \emph{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \emph{FedData} package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
Furthermore, the \emph{getlandsat} package \cite{Chamberlain2018} allows users to download Landsat 8 satellite data, the \emph{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \emph{sen2r} \cite{R-sen2r} to download Sentinel-2 optical images.
Also, the \texttt{getData()} function from the \emph{raster} package allows users to download climatic and bioclimatic data from WorldClim v1.4.
Additionally, the \emph{rgee} package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.

@@

Coordinate references systems (CRS) describe how spatial data is projected from the earth's three-dimensional surface to a two-dimensional surface as required for spatial analysis or creating maps .
This is also referred to as the spatial projection and is often the first barrier in spatial data analysis.
It is not only required to have all of the used data in the same projection, but also to select a proper CRS.
This is of importance because the projection into a two-dimensional surface unavoidable leads to distortion, and different CRS are optimized for different properties, regions of the world, and scales .
Coordinates in spatial data represent one of many coordinate reference systems.
Two main groups of CRS, namely geographical and projected, exist, with each having many members.
Using geographical CRS, positions are specified by latitude and longitude coordinates in degrees.
However, most landscape ecology studies should utilize projected CRSs, which use some measurement units (e.g., meters).
The selection of projected CRS should be based on the property of spatial data that needs to be kept intact (e.g., no distortion of areas, shapes, distances, or angles) and be appropriate for a given study area.
A common way to refer to different CRS is to used codes developed by the European Petrol Survey Group (EPSG).
Tools to find an appropriate CRS for a certain region can be found at \url{https://spatialreference.org}, \url{https://epsg.org}, or \url{http://epsg.io}.
All packages from Section \ref{sec:spatrep} have interfaces for coordinates transformations, allowing unification of spatial projections when the used data have different CRS.

@@

Finally, there are a number of tools for landscape ecology implemented in GIS software, such as \texttt{r.li} or \texttt{r.pi} for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.
Gladly, it is possible to control several GIS software directly from R using dedicated packages, such as \emph{rgrass7} \cite{R-grass7} for GRASS GIS, \emph{RSAGA} \cite{R-RSAGA} for SAGA GIS, and \emph{rgee} \cite{R-rgee} for Google Earth Engine.

@@

The \emph{tmap} package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \emph{cartography} package \cite{Giraud2016}.
Also the popular plotting package \emph{ggplot2} \cite{Wickham2016a} has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{R-ggspatial}.

Quick interactive visualization of spatial data can be done with the \emph{mapview} package\cite{R-mapview}.
Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{R-leaflet}.

A slightly different approach to visualizing spatial data is adapted by the \emph{rayshader} package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.

\begin{figure}
\includegraphics[width=1\linewidth,height=0.5\textheight]{data/plot_all} \caption{Comparison of different options to create maps using a) base plot, b) ggplot2, and c) tmap. All maps show the total annual precipitation of Switzerland. Data was downloaded with the raster (precipitation data) and rnaturalearth (country borders  package. The code to create the maps can be found in the appendix.}\label{fig:fig-map}
\end{figure}

@@

Related to SDMs, there are a growing number of \emph{R} packages to analyze data from tracked animals, to study their movement characteristics, space use, and interaction with other animals and the environment.
These analyses often use results of landscape ecological analyses as predictor variables to explain variation in space use \cite{signer2015}, behavioral states \cite{langrock2012}, or habitat selection \cite{fieberg2020}.
Widely used \emph{R} packages include \emph{ctmm} \cite{calabrese2016} and \emph{adehabitatHR} \cite{Calenge2006} for home-range estimation, \emph{moveHMM} \cite{michelot2016} for the classification of behavioral states, and \emph{amt} \cite{Signer2019} for habitat selection.
A recent and very comprehensive overview of R packages for the analysis of animal movement data is given by \cite{joo2020}.

@@

We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10\% of participants included ``biodiversity'', followed by ``land use management'', ``landscape connectivity'', and ``nature conservation'' (Fig. \ref{fig:fig-survey} A)).

Next, we were interested in the most important tasks to the workflow of the participants.
Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'', and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B).
The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').

@@

\caption{Results of the online survey about open-source software tools in R for landscape ecology. Results include A) which terms describe major reserach topics the best, B) the most important workflow task, C) the most used R packages and D) the overall usefulnes or R for landscape ecology. The 'others' category includes all answers with less than five total mentions.}\label{fig:fig-survey}
\end{figure}

@@

A comprehensive collection of \emph{R} software packages exist to handle the most common tasks and fields of landscape ecology.
Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, \emph{R} allows for transparent and reproducible workflows.
Additionally, this also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.

The survey revealed that overall the landscape ecology community is satisfied with the capabilities of the \emph{R} programming language for landscape ecology.
Furthermore, the survey showed the highly dynamic development of \emph{R} packages by the community itself, showing that avaiable tools are constantly updated.

Landscape ecology combines and covers many different research topics and methodological approaches and most of them have in common that they heavily rely on spatial data.
While the \emph{R} programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands in terms of computational time as well as memory requirements the \emph{R} programming language has to face.
Furthermore, as the individual fields collected under the umbrella of landscape ecology develop, so need the \emph{R} packages related to those fields.

>>>>

Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g.~Python) for landscape ecology exist, in this review we focus on software implemented in the \emph{R} programming language.
For more general overviews of open-source software for landscape ecology see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.
The \emph{R} programming language can be a very powerful tool. In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the \emph{R} environment (Fig. \ref{fig:fig-workflow}).
The growing body of \emph{R} packages related to spatial data processing and analysis results in a high capability of this language for landscape ecology.

@@

The guaranteed technical quality standard of \emph{CRAN} requires more initial work for developers compared to \emph{GitHub}, while it ensures for users that the package can be installed on their machine.
Additionally, the technical quality standard of \emph{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \emph{R} packages, i.e., package \emph{x} requires and uses code from package \emph{y}.
CRAN also provides archived versions of outdated or orphaned packages and thus ensuring long term availability and reproducibility of code.
Thus, most packages can be found on both platforms, and many developers use \emph{GitHub} for regular development and \emph{CRAN} to publish stable releases of the packages.
Furthermore, online communities like \emph{rOpenSci} also provide a peer-review process for code quality.
However, while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backwards compatibility is not always ensured.
Packages that deal with such issue include \emph{groundhog} \cite{Simonsohn2021}, \emph{packrat} \cite{Ushey2018}, or \emph{renv} \cite{Ushey2020}.
For more information about \emph{R} package development in general, see \cite{Wickham2015}.

@@

The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
In many cases, this data model allows a more realistic representation of landscape features, however, with the cost of higher computational demands \cite{Lovelace2019}.
The \emph{sp} package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 \emph{R} packages directly depend or imports \emph{sp}.
However, \emph{sp} is not actively developed anymore, and its recommended successor is the \emph{sf} package \cite{Pebesma2018}.
Besides many advantages and strength of the \emph{sf} package in terms of spatial data handling, it also integrates into the widely used \emph{tidyverse} packages \cite{Wickham2019a}.
\emph{sf} builds on the idea of ``simple features,'' a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points) and attributes connected to these geometries \cite{Pebesma2019a}.

@@

Nowadays, spatial data at many scales is available from an abundance of online-accessible sources.
A lot of this data are publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an \emph{R} session.
Since publicly available data is becoming more prominent, so are \emph{R} packages to access them.
Packages include \emph{rnaturalearth} \cite{South2017} to access the Natural Earth database to download region and country data, the \emph{elevatr} package to access raster elevation data \cite{R-elevatr}, the \emph{rgbif} package to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the \emph{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \emph{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \emph{FedData} package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
Furthermore, the \emph{getlandsat} package \cite{Chamberlain2018} allows users to download Landsat 8 satellite data, the \emph{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \emph{sen2r} \cite{R-sen2r} to download Sentinel-2 optical images.
Also, the \texttt{getData()} function from the \emph{raster} package allows users to download climatic and bioclimatic data from WorldClim v1.4.
Additionally, the \emph{rgee} package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.

@@

Coordinate references systems (CRS) describe how spatial data is projected from the earth's three-dimensional surface to a two-dimensional surface as required for spatial analysis or creating maps \cite{Lovelace2019,Pebesma2019a}.
This is also referred to as the spatial projection and is often the first barrier in spatial data analysis.
It is not only required to have all of the used data in the same projection, but also to select a proper CRS.
This is of importance because the projection into a two-dimensional surface unavoidable leads to distortion, and different CRS are optimized for different properties, regions of the world, and scales \cite{Bivand2013,Lovelace2019}.
Coordinates in spatial data represent one of many coordinate reference systems.
Two main groups of CRS, namely geographical and projected, exist, with each having many members.
Using geographical CRS, positions are specified by latitude and longitude coordinates in degrees.
However, most landscape ecology studies should utilize projected CRSs, which use some measurement units (e.g., meters).
The selection of projected CRS should be based on the property of spatial data that needs to be kept intact (e.g., no distortion of areas, shapes, distances, or angles) and be appropriate for a given study area.
A common way to refer to different CRS is to used codes developed by the European Petrol Survey Group (EPSG).
Tools to find an appropriate CRS for a certain region can be found at \url{https://spatialreference.org}, \url{https://epsg.org}, or \url{http://epsg.io}.
All packages from Section \ref{sec:spatrep} have interfaces for coordinates transformations, allowing unification of spatial projections when the used data have different CRS.

@@

Finally, there are a number of tools for landscape ecology implemented in GIS software, such as \texttt{r.li} or \texttt{r.pi} for GRASS GIS \cite{wegmann2018r,neteler2012grass,Porta2017}, terrain analysis methods in SAGA GIS \cite{gmd-8-1991-2015}, or morphological operations for Google Earth Engine.
It is possible to control several GIS software directly from \emph{R} using dedicated packages, such as \emph{rgrass7} \cite{R-grass7} for GRASS GIS, \emph{RSAGA} \cite{R-RSAGA} for SAGA GIS, and \emph{rgee} \cite{R-rgee} for Google Earth Engine.

@@

Further plotting methods for raster objects can be found in the \emph{rasterVis} package \cite{Lamigueiro2020}.
Also the popular plotting package \emph{ggplot2} \cite{Wickham2016a} has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{R-ggspatial}.
Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \emph{cartography} package \cite{Giraud2016}.
The \emph{tmap} package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
Quick interactive visualization of spatial data can be done with the \emph{mapview} package\cite{R-mapview}.
Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{R-leaflet}.
A slightly different approach to visualizing spatial data is adapted by the \emph{rayshader} package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.

\begin{figure}
\includegraphics[width=1\linewidth,height=0.5\textheight]{data/plot_all} \caption{Comparison of different options to create maps using a) base plot, b) ggplot2, and c) tmap. All maps show the total annual precipitation of Switzerland. Data was downloaded with the raster (precipitation data) and rnaturalearth (country borders) package. The code to create the maps can be found in the Appendix.}\label{fig:fig-map}
\end{figure}

@@

Related to SDMs, there is a growing number of \emph{R} packages to analyze data from tracked animals, to study their movement characteristics, space use, and interaction with other animals and the environment.
These analyses often use results of landscape ecological analyses as predictor variables to explain variation in space use \cite{signer2015}, behavioral states \cite{langrock2012}, or habitat selection \cite{fieberg2020}.
Widely used \emph{R} packages include \emph{ctmm} \cite{calabrese2016} and \emph{adehabitatHR} \cite{Calenge2006} for home-range estimation, \emph{moveHMM} \cite{michelot2016} for the classification of behavioral states, and \emph{amt} \cite{Signer2019} for habitat selection.
A recent and very comprehensive overview of R packages for the analysis of animal movement data is given by \cite{joo2020}.

@@

We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10\% of participants included ``biodiversity,'' followed by ``land use management,'' ``landscape connectivity,'' and ``nature conservation'' (Fig. \ref{fig:fig-survey} A)).

Next, we were interested in the most important tasks to the workflow of the participants.
Not surprisingly, ``(pre-)processing of data,'' ``spatial statistics,'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B).
The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').

@@

\caption{Results of the online survey about open-source software tools in R for landscape ecology. Results include a) which terms describe major reserach topics the best, b) the most important workflow task, c) the most used R packages and d) the overall usefulnes or R for landscape ecology. The 'others' category includes all answers with less than five total mentions.}\label{fig:fig-survey}
\end{figure}

@@

A comprehensive collection of \emph{R} software packages exist to handle the most common tasks of landscape ecology.
Because it is possible to import, modify, analyze, and visualize spatial data all in the same programming environment, \emph{R} allows for transparent and reproducible workflows.
Additionally, this also allows users to easily interchange, modify, or adapt methods from other related and unrelated fields.

The survey revealed that the landscape ecology community is overall satisfied with the capabilities of the \emph{R} programming language for landscape ecology.
Furthermore, the survey showed the that many members of the landscape ecology community actively develop \emph{R} packages themselves, demonstrating that tools are constantly added and updated.

Landscape ecology combines many different research topics and methodological approaches and most of them heavily rely on spatial data.
While the \emph{R} programming language is generally well suited to handle, analyze and visualize spatial data, the increasing availability of large data sets also leads to the challenges of increased computational demands, in terms of computational time as well as memory requirements, the \emph{R} programming language has to face.
Furthermore, as the individual fields collected under the umbrella of landscape ecology develop, so need the \emph{R} packages related to those fields.
But as the development of \emph{R} packages over the past years clearly demonstrated, the landscape ecology community is ready to face these challenges.
